Evolution of the AGN UV luminosity function

### 2017-09-18 Paper/paper.tex
Update text

  In each bin, we estimate the uncertainty in the luminosity function
  by assuming Poisson statistics \citep{1986ApJ...303..336G}.  The
  resultant binned luminosity function estimates are shown by the
  points in Figures \ref{fig:mosaic} and \ref{fig:mosaic2}.  These
  agree quite well with the published luminosity functions by these
  surveys.

@@

@@

As our next step, we stop binning the quasars in luminosity, and only
bin them in redshifts.  In each redshift bin, we model the quasar
luminosity function as a double power law given by
\citep{1988MNRAS.235..935B, 1995ApJ...438..623P, 2000MNRAS.317.1014B}
\begin{equation}
  \phi(M) =
  \frac{\phi_*}{10^{0.4(\alpha+1)(M-M_*)}+10^{0.4(\beta+1)(M-M_*)}}.
  \label{eqn:dpl}
\end{equation}
Equation~(\ref{eqn:dpl}) has four free parameters: the amplitude
$\phi_*$, the ``break'' $M_*$, the faint-end slope $\beta$, and the
bright-end slope $\alpha$.  We obtain posterior probability
distributions for these parameters using MCMC \citep[e.g.,][]{jaynes}.
The joint posterior probability distribution of the model parameters
is then written as
\begin{multline}
  p(\phi_*, M_*, \alpha, \beta | \{M_i, z_i\}) \propto \\ p(\phi_*, M_*, \alpha, \beta)p(\{M_i, z_i\} | \phi_*, M_*, \alpha, \beta),
\end{multline}
where the constant of proportionality is independent of the luminosity
function parameters, and $\{M_i, z_i\}$ denotes the complete quasar
data.  We use a uniform prior distribution $p(\phi_*, M_*, \alpha,
\beta)$ and assume that the likelihood
\begin{equation}
  \mathcal{L}\equiv p(\{M_i, z_i\} | \phi_*, M_*, \alpha, \beta)
\end{equation}
is given by $\phi(M)$ with suitable normalisation.  That is the
negative logarithm of the likelihood $S\equiv -2\ln\mathcal{L}$ can be
written as
\begin{multline}
  S = -2\sum_{i=1}^N\ln\phi(M_i, z_i)\\+2\int_{M_\mathrm{min}}^{M_\mathrm{max}}dM\int_{z_\mathrm{min}}^{z_\mathrm{max}}dz\, \phi(M,z) f(M, z)\,\frac{dV}{dz}.
  \label{eqn:S}
\end{multline}
where $N$ is the total number of quasars in the sample and the
integral in the second term on the right hand side is on the 
surveyed range of $M$ and $z$.

Another way of understanding the above likelihood is as limit of the
Poisson likelihood in luminosity and redshift bins
\citep{1983ApJ...269...35M, 2001AJ....121...54F}.  We can write the
probability of observing $n_{ij}$ quasars in the $(M_i, z_j)$ bin as
the Poisson distribution
\begin{equation}
  \mathcal{L}=\prod_{i,j}\frac{e^{-\mu_{ij}}\mu_{ij}^{n_{ij}}}{n_{ij}!},
  \label{eqn:lhood}
\end{equation}
where 
\begin{equation}
  \mu_{ij}= \int_{M_i}^{M_{i+1}}dM\int_{z_j}^{z_{j+1}}dz\, \phi(M,z) f(M, z)\,\frac{dV}{dz},
\end{equation}
is the average number of quasars expected in the $(M_i, z_j)$ bin
given the luminosity function $\phi(M,z)$.  In the limit of
infinitesimal bins, $n_{ij}=0$ or $1$, in which case
Equation~(\ref{eqn:lhood}) can be simplified and we can get
Equation~(\ref{eqn:S}).

Our estimates for the luminosity function in several redshift bins are
shown in Figures \ref{fig:mosaic} and \ref{fig:mosaic2}.  The median
of the posterior is shown as the best fit model.  Yellow curves show a
random draw from the posterior.

@@

It is useful to go beyond the redshift bins of the previous section.
This gives us continuous models of luminosity function evolution and
is also relevant for extrapolation. We therefore obtain estimates of
the luminosity function.

@@

\begin{figure*}
  \begin{center}
    % \includegraphics[width=\textwidth,keepaspectratio]{mosaic_small.pdf}
    \includegraphics[height=0.9\textheight,keepaspectratio]{mosaic_small.pdf}
  \end{center}
  \caption{Individual luminosity function fits for $z=0.3$ to $2.6$.}
  \label{fig:mosaic}
\end{figure*}

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth]{evolution_alldata.pdf}
  \end{center}
  \caption{Parameter evolution from individual fits in
    Figures~\ref{fig:mosaic} and \ref{fig:mosaic2}.  Parameters from
    BOSS quasars are seen to be systematically offset. \textbf{[Show
        $z < 0.6$ point also?]}}
\end{figure*}

>>>>

  In each bin, we estimate the uncertainty in the luminosity function
  by assuming Poisson statistics \citep{1986ApJ...303..336G}.  The
  resultant binned luminosity function estimates are shown by the
  points in Figure~\ref{fig:mosaic}.  These agree quite well with the
  published luminosity functions by these surveys.

@@

\begin{figure*}
  \begin{center}
    % \includegraphics[width=\textwidth,keepaspectratio]{mosaic_small.pdf}
    \includegraphics[height=0.88\textheight,keepaspectratio]{mosaic_small.pdf}
  \end{center}
  \caption{Luminosity function estimates from $z=0.3$ to $2.6$.  The
    symbols show our inferred binned luminosity functions.  In each
    redshift bin, the black curve shows the best-fit double power law
    model, which is represented by the median of the posterior
    probability distribution function.  Yellow curves show a random
    draw from the posterior probability distribution function.  As
    discussed in the text, luminosity bins that show a decrease in the
    luminosity function at the faint end are deemed to be incomplete
    are are removed from our analysis while fitting the double power
    law model.  These bins are represented by open symbols in this
    figure.  Each panel shows the number of quasars in the respective
    redshift bin.  The number in parenthesis indicates the total
    number of quasars in the bin, before excluding incomplete
    bins.}
  \label{fig:mosaic}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics[width=\textwidth]{evolution_alldata.pdf}
  \end{center}
  \caption{Parameter evolution from individual fits in
    Figures~\ref{fig:mosaic}.  Parameters from BOSS quasars are seen
    to be systematically offset.}
  \label{fig:evoln}
\end{figure*}

@@

As our next step, we stop binning the quasars in luminosity, and only
bin them in redshifts.  In each redshift bin, we model the quasar
luminosity function as a double power law given by
\citep{1988MNRAS.235..935B, 1995ApJ...438..623P, 2000MNRAS.317.1014B}
\begin{equation}
  \phi(M) =
  \frac{\phi_*}{10^{0.4(\alpha+1)(M-M_*)}+10^{0.4(\beta+1)(M-M_*)}}.
  \label{eqn:dpl}
\end{equation}
Equation~(\ref{eqn:dpl}) has four free parameters: the amplitude
$\phi_*$, the ``break'' $M_*$, the faint-end slope $\beta$, and the
bright-end slope $\alpha$.  We obtain posterior probability
distributions for these parameters using MCMC \citep[e.g.,][]{jaynes}.
The joint posterior probability distribution of the model parameters
is then written as
\begin{multline}
  p(\phi_*, M_*, \alpha, \beta | \{M_i, z_i\}) \propto \\ p(\phi_*, M_*, \alpha, \beta)p(\{M_i, z_i\} | \phi_*, M_*, \alpha, \beta),
\end{multline}
where the constant of proportionality is independent of the luminosity
function parameters, and $\{M_i, z_i\}$ denotes the magnitudes and
redshifts of quasars falling in a redshift bin $[z_\mathrm{min},
  z_\mathrm{max})$.  We use a uniform prior distribution $p(\phi_*,
  M_*, \alpha, \beta)$ and assume that the likelihood
\begin{equation}
  \mathcal{L}\equiv p(\{M_i, z_i\} | \phi_*, M_*, \alpha, \beta)
\end{equation}
is given by $\phi(M)$ with suitable normalisation.  The negative
logarithm of the likelihood $S\equiv -2\ln\mathcal{L}$ can then be
written as
\begin{multline}
  S = -2\sum_{i=1}^N\ln\phi(M_i, z_i)\\+2\int_{M_\mathrm{min}}^{M_\mathrm{max}}dM\int_{z_\mathrm{min}}^{z_\mathrm{max}}dz\, \phi(M,z) f(M, z)\,\frac{dV}{dz}.
  \label{eqn:S}
\end{multline}
where $N$ is the total number of quasars in the redshift bin and the
luminosity integral in the second term on the right hand side is on
the surveyed range of $M$.

The above likelihood can also be understood as the limit of the
Poisson likelihood in luminosity and redshift bins
\citep{1983ApJ...269...35M, 2001AJ....121...54F}.  We can write the
probability of observing $n_{ij}$ quasars in the $(M_i, z_j)$ bin as
the Poisson distribution
\begin{equation}
  \mathcal{L}=\prod_{i,j}\frac{e^{-\mu_{ij}}\mu_{ij}^{n_{ij}}}{n_{ij}!},
  \label{eqn:lhood}
\end{equation}
where 
\begin{equation}
  \mu_{ij}= \int_{M_i}^{M_{i+1}}dM\int_{z_j}^{z_{j+1}}dz\, \phi(M,z) f(M, z)\,\frac{dV}{dz},
\end{equation}
is the average number of quasars expected in the $(M_i, z_j)$ bin
given the luminosity function $\phi(M,z)$.  In the limit of
infinitesimal bins, $n_{ij}=0$ or $1$, Equation~(\ref{eqn:lhood}) can
be simplified and we get Equation~(\ref{eqn:S}).

Our estimates for the luminosity function in several redshift bins are
shown in Figure~\ref{fig:mosaic}.  The median of the posterior is
shown as the best fit model.  Yellow curves show a random draw from
the posterior.  Consistent with previous studies, the double power law
model provides an excellent description of the luminosity function
model over almost the complete range of redshifts spanned by the data.
It is only in the highest redshift bin ($z=5.5$--$6.5$) that the data
seem to favour a single power law.  In this bin, we the resultant
posterior distribution of the break magnitude $M_*$ is bimodal with
favoured values at the faint ($M_*>-18$) and bright end of the data
($M_*<-30$).  We consider only the latter possibility as consistent
with the data at lower redshifts ($z<5.5$).  In light of the low
redshift data, we use restricted priors in this redshift bins: we
restrict the bright-end slope $\alpha$ to values less than $-4$.
Other paramters continue to have wide uniform priors.

As seen in Figure~\ref{fig:mosaic}, in several redshift bins the
binned quasar luminosity function seems to show a decline at the faint
end.  This is clearly seen, for example, in the 2SLAQ sample at $z<2$.
We take this to be a systematic error: the decrease in the luminosity
function is an indication of incompleteness in the quasar surveys that
has not been accounted for by the reported completeness map.  We
choose to discard such quasars from our analysis.  This behaviour is
also seen in the SDSS quasar sample at $z<1.8$ and $z\sim 4$.  The
discarded luminosity bins are shown in Figure~\ref{fig:mosaic} by open
circles.  We can see that the surviving quasar sample is in excellent
agreement with the double power law model at all redshifts.

The values of the four double power law parameters in various redshift
bins are shown in Figure~\ref{fig:evoln}.  We find interesting
evolutionary trends in each of the four parameters.  The break
luminosity $M_*$ evolves by more than eight magnitudes from redshift
$z=0$ to $7$.  The amplitude of the luminosity function $\phi_*$
evolves moderately from $z=0$ to $z\sim 3$ and then drops by six
orders of magnitude to about $10^{-12}$ at $z\sim 7$.  The bright end
slope $\alpha$ has significant scatter, but still shows a trend
towards larger negative values---that is, steeper bright-end
luminosity functions---at high redshifts.  Finally, the faint end of
the luminosity function also shows signs of increasing steepness
towards high redshifts.  Somewhat similar to the amplitude $\phi_*$,
the faint-end slope $\beta$ also shows a break at $z\sim 3$.  It stays
roughly constant up to this redshift, and the drops to increasingly
negative values at high redshifts, indicating a steeper faint end.

The behaviour of BOSS quasars in Figure~\ref{fig:evoln} is striking.
In the redshift range covered by BOSS the values of the four double
power law parameters deviate from the general trends.  The faint end
of the luminosity function of the BOSS quasars is much shallower and
the break luminosity is fainter by a magnitude.  The bright end of the
luminosity function of quasars in this sample is also much flatter
than that for other samples.  The conspicuousness of the deviation of
the BOSS sample from the general trends makes it unlikely that the
luminosity function evolution indicated by BOSS in the redshift range
$z=2$--$3$ is physical.  Indeed, in this redshift range BOSS quasar
selection is known to be affected by a colour bias
\citep{2011ApJ...728...23W}, as the quasar colour selection is
contaminated by a population of blue stars.  We therefore deem the
BOSS sample as affected by this systematic error.

The open circles in Figure~\ref{fig:evoln} show the results of a
similar analysis of the quasar UV luminosity function by
\citep{2017MNRAS.466.1160M}.  Evidently, our results are quite
different from the results of these authors.  While their analysis
generally also reveals a steep faint end slop, the break luminosity
evolution is dramatically different.  In their analysis the break
luminosity tends to remain roughly constant $M_*\sim -24$ from $z=0$
to $7$, but this does not seem to be indicated by the data shown in
Figure~\ref{fig:mosaic}.  A possible reason behind their significanly
different results could be a bias introduced by binning in
luminosities and redshifts while fitting the luminosity function model.

@@

The smooth evolution of the luminosity function parameters apparant in
Figure~\ref{fig:evoln} suggests that it should be possible to describe
the quasar UV luminosity function evolution from $z\sim 7$ to $0$
using far fewer number of parameters.  Such descriptions have been put
forward in the literature for the X-ray and bolometric luminosity
functions \citep[e.g.,][]{2007ApJ...654..731H, 2015MNRAS.451.1892A}.
Such ``global'' descriptions of the luminosity function evolution are
useful as they give a continuous description of the luminosity
function.  This makes them valuable while developing an understanding
of the physics behind the luminosity function. We therefore obtain
estimates of the luminosity function in which the evolution of the
four parameters of the double power law models in
Equation~(\ref{eqn:dpl}) is parameterised to write
\begin{align}
  &\phi_*(z) = f_0(\{c_{0,i}\}, z)\nonumber\\
  &M_*(z) = f_1(\{c_{1,i}\}, z)\nonumber\\
  &\alpha(z) = f_2(\{c_{2,i}\}, z)\nonumber\\
  &\beta(z) = f_3(\{c_{3,i}\}, z),
\end{align}
where the $c_i$'s are the new model parameters.  The joint posterior
probability distribution of these parameters can be now written as
\begin{equation}
  p(\{c_i\} | \{M_i, z_i\}) \propto p(\{c_i\})p(\{M_i, z_i\} | \{c_i\}),
\end{equation}
where the likelihood 
\begin{equation}
  \mathcal{L}\equiv p(\{M_i, z_i\} | \{c_i\}),
\end{equation}
is now given by $\phi(M,z)$ with suitable normalisation.  Note that
$\phi(M,z)$ is simply 
\begin{multline}
  \phi(M,z) = \phi_*(z) \left[10^{0.4(\alpha(z)+1)(M-M_*(z))}\right. \\ \left.+ 10^{0.4(\beta(z)+1)(M-M_*(z))}\right]^{-1}.
\end{multline}
The ngative logarithm of the likelihood $S\equiv -2\ln\mathcal{L}$ is
straightforward generalisation of Equation~(\ref{eqn:S}), so that
\begin{multline}
  S = -2\sum_{i=1}^N\ln\phi(M_i, z_i)\\+2\int_{M_\mathrm{min}}^{M_\mathrm{max}}dM\int_{z_\mathrm{min}}^{z_\mathrm{max}}dz\, \phi(M,z) f(M, z)\,\frac{dV}{dz},
  \label{eqn:S2}
\end{multline}
where now $N$ is the total number of quasars included in the analysis,
and the integrals in the second term on the right-hand side are over
the complete surveyed range of luminosity and redshift.

@@

@@
